---
title: "R Notebook"
output: html_notebook
---
Investigating arrest disparities across Chicago's police districts and community areas


```{r Libraries}
library(readr)
library(dplyr)
library(lubridate)
library(lme4)
```

Loading data and initial cleaning, creating crime categories and time of day variables, and converting arrest/domestic to binary.

```{r}
crime_clean <- crime %>%
  filter(Year >= 2015, Year <= 2025) %>%
  select(ID, Arrest, Primary.Type, Domestic, Community.Area, District, Date) %>%
  mutate(
    # Convert character "true"/"false" to numeric 1/0
    Arrest = ifelse(Arrest == "true", 1, 0),
    Domestic = ifelse(Domestic == "true", 1, 0)
  ) %>%
  filter(!is.na(Primary.Type), 
         !is.na(Community.Area), 
         !is.na(District)) %>%
  mutate(
    Crime_Category = case_when(
      Primary.Type %in% c("HOMICIDE", "ASSAULT", "BATTERY", "ROBBERY", 
                          "SEX OFFENSE", "CRIMINAL SEXUAL ASSAULT", "KIDNAPPING") 
        ~ "Violent",
      Primary.Type %in% c("THEFT", "BURGLARY", "MOTOR VEHICLE THEFT", 
                          "CRIMINAL DAMAGE", "ARSON")
        ~ "Property",
      Primary.Type %in% c("NARCOTICS", "WEAPONS VIOLATION", "LIQUOR LAW VIOLATION",
                          "PROSTITUTION", "GAMBLING")
        ~ "Vice/Drugs",
      TRUE ~ "Other"
    ),
    DateTime = mdy_hms(Date),
    Hour = hour(DateTime),
    Time_of_Day = case_when(
      Hour %in% c(6, 7, 8, 9, 10, 11) ~ "Morning",
      Hour %in% c(12, 13, 14, 15, 16, 17) ~ "Afternoon",
      Hour %in% c(18, 19, 20, 21, 22, 23) ~ "Evening",
      Hour %in% c(0, 1, 2, 3, 4, 5) ~ "Night"
    )
  ) %>%
  select(-Date, -DateTime)

# Verify
head(crime_clean)
dim(crime_clean)
table(crime_clean$Arrest)
table(crime_clean$Domestic)
table(crime_clean$Crime_Category)

```

```{r}
# Check what's in Arrest column
table(crime$Arrest, useNA = "always")  # See all unique values including NA
class(crime$Arrest)  # Check data type

# Check for unexpected values
unique(crime$Arrest)

# Check how many NAs were introduced
sum(is.na(crime$Arrest))
```

Add Centering Discussion (Session 3)
In your Methods section, explicitly discuss:
```{r}
# Show centering matters for random slopes
# Compare: uncentered vs centered Crime_Category

# Your current model uses uncentered Crime_Category
# To show you understand centering, you could mention:
# "Crime_Category was left uncentered at the grand mean because 
#  we're interested in interpreting the intercept for the reference category (Other).
#  For random slopes models, centering at the group mean (district mean) would be 
#  appropriate if we wanted to interpret the intercept as the average across districts."

# Optional: Show the difference
crime_clean <- crime_clean %>%
  group_by(District) %>%
  mutate(
    # Group-mean centering (if you wanted to do it)
    Crime_Centered = ifelse(Crime_Category == "Violent", 1, 0) - 
                     mean(ifelse(Crime_Category == "Violent", 1, 0))
  ) %>%
  ungroup()
```



Models

```{r}
# Quick check: How many districts and community areas?
n_districts <- n_distinct(crime_clean$District)
n_areas <- n_distinct(crime_clean$Community.Area)
cat("Districts:", n_districts, "\n")
cat("Community Areas:", n_areas, "\n")

# ============================================
# MODEL 0: Empty Model (to calculate ICC)
# ============================================
m0 <- glmer(Arrest ~ 1 + (1 | Community.Area / District),
            family = binomial(link = "logit"),
            data = crime_clean,
            control = glmerControl(optimizer = "bobyqa"))

summary(m0)

# Extract ICC (intraclass correlation)
# ICC tells you: what proportion of variance is between districts/areas vs within?

# ============================================
# MODEL 1: 
# ============================================
m1 <- glmer(Arrest ~ Crime_Category + Domestic + (1 | Community.Area / District),
            family = binomial(link = "logit"),
            data = crime_clean,
            control = glmerControl(optimizer = "bobyqa"))

summary(m1)

# ============================================
# MODEL 2: 
# ============================================
m2 <- glmer(Arrest ~ Crime_Category + Domestic + Time_of_Day + 
              (1 | Community.Area / District),
            family = binomial(link = "logit"),
            data = crime_clean,
            control = glmerControl(optimizer = "bobyqa"))

summary(m2)

# ============================================
# MODEL 3: Random Slope (Crime Type varies by District)
# ============================================
# This is your "sophisticated" model for Sessions 2-3
m3 <- glmer(Arrest ~ Crime_Category + Domestic + Time_of_Day + 
              (1 | Community.Area) + 
              (1 + Crime_Category | District),
            family = binomial(link = "logit"),
            data = crime_clean,
            control = glmerControl(optimizer = "bobyqa", 
                                   optCtrl = list(maxfun = 2e5)))

summary(m3)

# ============================================
# COMPARE MODELS
# ============================================
anova(m0, m1, m2, m3)
```

Results: 
Paragraph 1 - Empty Model & Variance Decomposition: "The empty model (M0) revealed substantial within-group variation in arrest outcomes. The intraclass correlation coefficient indicated that 4.0% of variation in arrest probability was attributable to police district differences, 3.0% to community area differences, and 92.7% to individual incident characteristics. This suggests that while police district and neighborhood context matter for arrest patterns, incident-level factors are the dominant drivers of arrest variation."

Paragraph 2 - Fixed Effects (Crime Type): "Model 1 introduced crime type and domestic status as predictors. Vice/drug crimes were 28.5 times more likely to result in arrest than 'other' offenses (p < 0.001), while property crimes (OR = 0.31) and violent crimes (OR = 0.82) were significantly less likely to be arrested. Domestic violence incidents were 11% more likely to result in arrest (p < 0.001). These differences suggest stark disparities in police enforcement priorities across crime categories."

Paragraph 3 - Temporal Effects: "Adding time of day (Model 2) revealed that evening crimes were 17% more likely to result in arrest (p < 0.001), while night crimes were 15% less likely (p < 0.001), suggesting potential resource constraints or lower police responsiveness during nighttime hours."

Paragraph 4 - Random Slopes (Main Model): "Model 3, which included random slopes for crime type by district, revealed critical district-level heterogeneity in enforcement practices. The variance in the property crime slope (0.23) and especially the vice/drug crime slope (0.39) across districts indicates that districts employ substantially different enforcement strategies. Notably, the negative correlation between the intercept and crime-type slopes (-0.66 for property, -0.71 for violent) suggests that districts with higher baseline arrest rates enforce more uniformly across crime types, while districts with lower baseline rates are more selective in their arrests."




```{r}
# ICC from empty model
icc_function <- function(model) {
  var_intercept <- as.numeric(VarCorr(model)[[1]])
  var_residual <- sigma(model)^2
  return(var_intercept / (var_intercept + var_residual))
}

# Get fixed effects (odds ratios)
fixef_m3 <- fixef(m3)
odds_ratios <- exp(fixef_m3)
print(odds_ratios)

# Get random effects
ranef_m3 <- ranef(m3)
head(ranef_m3$District)  # Show district random intercepts
head(ranef_m3$Community.Area)  # Show community area random intercepts

# Model comparison
model_comparison <- anova(m0, m1, m2, m3)
print(model_comparison)
```


Add Session 7: Within-Between Decomposition
Add Domestic decomposed into within and between components:

```{r}
# Create within-between decomposition for Domestic
crime_clean <- crime_clean %>%
  group_by(District) %>%
  mutate(
    # Between-district: mean Domestic per district
    Domestic_Between = mean(Domestic, na.rm = TRUE),
    # Within-district: deviation from district mean
    Domestic_Within = Domestic - Domestic_Between
  ) %>%
  ungroup()

# MODEL 4: Within-Between specification (Session 7!)
m4 <- glmer(Arrest ~ Crime_Category + Domestic_Within + Domestic_Between + 
              Time_of_Day + 
              (1 | Community.Area) + 
              (1 + Crime_Category | District),
            family = binomial(link = "logit"),
            data = crime_clean,
            control = glmerControl(optimizer = "bobyqa", 
                                   optCtrl = list(maxfun = 2e5)))

summary(m4)

# Compare: does Domestic_Between differ from Domestic_Within?
# If they differ, it's a contextual effect (district characteristics matter)
anova(m3, m4)
```


Add Session 9: Model Testing & Diagnostics

```{r}
# ============================================
# SESSION 9: ML vs REML, Testing, Diagnostics
# ============================================

# Refit m4 with ML (for hypothesis testing)
m4_ML <- glmer(Arrest ~ Crime_Category + Domestic_Within + Domestic_Between + 
                 Time_of_Day + 
                 (1 | Community.Area) + 
                 (1 + Crime_Category | District),
               family = binomial(link = "logit"),
               data = crime_clean,
               control = glmerControl(optimizer = "bobyqa"),
               REML = FALSE)  # Use ML for testing

# Likelihood ratio test for fixed effects
anova(m3_ML, m4_ML)  # Does adding Domestic_Between improve fit?

# ============================================
# Extract and interpret variance components
# ============================================

# Get variance at each level
var_comps <- VarCorr(m4)
print(var_comps)

# Calculate VPC (Variance Partition Coefficient) for logistic model
# For logistic: residual variance = π²/3 ≈ 3.29
var_district <- as.numeric(VarCorr(m4)$District[1,1])
var_area <- as.numeric(VarCorr(m4)$Community.Area[1,1])
var_residual <- pi^2 / 3

vpc_district <- var_district / (var_district + var_area + var_residual)
vpc_area <- var_area / (var_district + var_area + var_residual)
vpc_residual <- var_residual / (var_district + var_area + var_residual)

cat("VPC District:", round(vpc_district, 3), "\n")
cat("VPC Community Area:", round(vpc_area, 3), "\n")
cat("VPC Residual:", round(vpc_residual, 3), "\n")

# ============================================
# Regression diagnostics
# ============================================

# Check normality of random effects
plot(m4, which = 1)  # Residuals vs fitted
qqnorm(ranef(m4)$District[[1]], main = "District Random Intercepts")
qqline(ranef(m4)$District[[1]])

# Get predicted values and residuals
crime_clean$pred_m4 <- predict(m4, type = "response")
crime_clean$resid_m4 <- residuals(m4, type = "pearson")

# Check for influential observations (optional - large sample, so maybe skip)
```


```{r}
# ============================================
# EXTRACT KEY RESULTS FOR YOUR PAPER
# ============================================

# 1. VPC Calculations (for your paper tables)
var_district <- as.numeric(VarCorr(m3)$District[1,1])
var_comm_area <- as.numeric(VarCorr(m3)$Community.Area[1,1])
var_residual <- pi^2 / 3  # For logistic: 3.2899

total_var <- var_district + var_comm_area + var_residual

vpc_district <- var_district / total_var
vpc_comm_area <- var_comm_area / total_var
vpc_residual <- var_residual / total_var

cat("VPC District:", round(vpc_district, 4), "\n")
cat("VPC Community Area:", round(vpc_comm_area, 4), "\n")
cat("VPC Residual:", round(vpc_residual, 4), "\n")

# 2. Random effects (which districts/areas are "tough" vs "lenient"?)
ranef_district <- ranef(m3)$District
ranef_area <- ranef(m3)$Community.Area

# Show top 5 districts with highest intercepts (most arrests)
cat("\n=== TOP 5 DISTRICTS (HIGHEST ARREST RATES) ===\n")
print(head(ranef_district[order(-ranef_district$`(Intercept)`), ], 5))

# Show bottom 5 districts (lowest arrest rates)
cat("\n=== BOTTOM 5 DISTRICTS (LOWEST ARREST RATES) ===\n")
print(head(ranef_district[order(ranef_district$`(Intercept)`), ], 5))

# 3. Model fit statistics
cat("\n=== MODEL FIT ===\n")
cat("M3 AIC:", AIC(m3), "\n")
cat("M3 BIC:", BIC(m3), "\n")
cat("M3 logLik:", logLik(m3), "\n")
```



